---
title: "R Notebook"
output: html_notebook
editor_options: 
  chunk_output_type: console
---
# Download PRISM data for historical

This code downloads and processes daily PRISM 800-meter resolution climate data for each National PArk (we just have to change the ecoregion input in the loop and run it for each park). PRISM (Parameter-elevation Regressions on Independent Slopes Model) is a high-resolution climate dataset produced by Oregon State University and provides gridded estimates of daily weather variables across the U.S. from 1981 to the present.

In this workflow, we:
	•	Download daily gridded PRISM data for four climate variables: minimum temperature (tmin), maximum temperature (tmax), precipitation (ppt), and dew point temperature (tdmean) using the official PRISM web service.
	•	Automatically retry failed downloads with exponential backoff and log successful retrievals to a file.
	•	Use a parallelized loop (future_lapply) to speed up processing across available CPU cores.
	•	Project and resample each downloaded raster to match a pre-defined ecoregion raster representing climate zones within Olympic National Park.
	•	Use zonal averaging to summarize the mean value of each climate variable by ecoregion for each day.
	•	Store the results in a single tidy CSV file with one row per ecoregion per day from 1981–2024.

The result is a complete, ecoregion-specific historical daily climate dataset ready for LANDIS-II.

```{r}
if (!require("pacman")) install.packages("pacman")
pacman::p_load("httr", "utils", "terra", "sf", "exactextractr", "future", "future.apply", "progressr", 'tidyverse', 'glue') # these are the libraries we are using
options(scipen=999) # no sci notation

setwd("/Volumes/BigStore/PNW_Parks/Spatial/")
setwd("/Users/jlamping/Dropbox (University of Oregon)/Lamping/NPS_postdoc/Spatial/")
# PRISM 800m Daily Downloader and Ecoregion Summarizer

# Set output directories
download_dir <- "Climate/prism_800m"
out_csv <- "Climate/outputs/prism_800m_daily_allparks.csv"
out_dir <- "Climate/outputs/prism_800m_daily"

# Function to download and unzip one day of PRISM data
download_prism_day <- function(variable, date, out_dir, retries = 3, timeout_sec = 60) {
  library(httr)
  library(glue)

  date <- as.Date(date)
  base_url <- "https://services.nacse.org/prism/data/get/us/800m"
  url <- paste0(base_url, "/", variable, "/", format(date, "%Y%m%d"))
  zip_path <- file.path(out_dir, paste0(variable, "_", format(date, "%Y%m%d"), ".zip"))

  # Skip if already downloaded
  if (!file.exists(zip_path)) {
    attempt <- 1
    success <- FALSE

    while (attempt <= retries && !success) {
      resp <- try(
        GET(url, write_disk(zip_path, overwrite = TRUE), timeout(timeout_sec)),
        silent = TRUE
      )
      success <- inherits(resp, "response") && resp$status_code == 200

      if (!success) {
        message(glue("⏳ Retry {attempt}/{retries} for {url}..."))
        Sys.sleep(3 * attempt)  # exponential backoff
        attempt <- attempt + 1
      }
    }

    if (!success) {
      message("❌ Failed to download after retries: ", url)
      return(NULL)
    }
  }

  unzip_dir <- file.path(out_dir, paste0(variable, "_", format(date, "%Y%m%d")))
  if (!dir.exists(unzip_dir)) {
    try(unzip(zip_path, exdir = unzip_dir), silent = TRUE)
  }

  # ✅ Log success
  log_dir <- "outputs/logs"
  dir.create(log_dir, recursive = TRUE, showWarnings = FALSE)
  write(format(date, "%Y-%m-%d"), file = file.path(log_dir, "downloaded_dates.txt"), append = TRUE)

  return(unzip_dir)
}

# Define variables and date range
vars <- c("tmin", "tmax", "ppt", "tdmean")

# Include full year
dates <- seq(as.Date("1981-01-01"), as.Date("2024-12-31"), by = "day")

# Parallel plan
plan(multisession, workers = parallel::detectCores() - 1)

with_progress({
  p <- progressor(along = dates)

  all_results <- future_lapply(dates, function(date) {

    library(terra)
    library(httr)
    library(utils)

    # Load ecoregion raster inside worker
    ecoregion_raster <- rast("Ecoregions/Raster/MORA_ClimateRegions.tif")

    day_stack <- list()

    for (v in vars) {

      folder <- download_prism_day(v, as.Date(date), out_dir = download_dir)
      if (is.null(folder)) return(NULL)

      tif <- list.files(folder, pattern = "\\.tif$", full.names = TRUE)
      if (length(tif) == 0 || !file.exists(tif[1])) return(NULL)

      r <- rast(tif)

      # ==========================================================
      # FIX 1: Properly treat PRISM NoData before interpolation
      # ==========================================================
      # PRISM commonly uses -9999 as NoData
      try(terra::NAflag(r) <- -9999, silent = TRUE)

      # For precipitation, any negative value is invalid.
      # Remove BEFORE resampling to prevent bilinear bleed.
      if (v == "ppt") {
        r[r < 0] <- NA
      }

      # Project and resample
      r <- project(r, crs(ecoregion_raster))
      r <- resample(r, ecoregion_raster, method = "bilinear")

      # ==========================================================
      # FIX 2: Safety clamp for ppt after resampling
      # ==========================================================
      if (v == "ppt") {
        r[r < 0] <- 0
      }

      names(r) <- v
      day_stack[[v]] <- r
    }

    if (length(day_stack) == length(vars)) {

      stack <- rast(day_stack)

      zonal_stats <- as.data.frame(
        terra::zonal(stack, ecoregion_raster,
                     fun = "mean", na.rm = TRUE)
      )

      names(zonal_stats)[1] <- "ecoregion"
      zonal_stats$date <- date

      # Cleanup downloaded files
      for (v in vars) {
        folder <- file.path(download_dir,
                            paste0(v, "_", format(as.Date(date), "%Y%m%d")))
        zip_path <- file.path(download_dir,
                              paste0(v, "_", format(as.Date(date), "%Y%m%d"), ".zip"))

        try(unlink(folder, recursive = TRUE), silent = TRUE)
        try(unlink(zip_path), silent = TRUE)
      }

      p()
      return(zonal_stats)

    } else {
      p()
      return(NULL)
    }

  }, future.seed = TRUE)
})

climate_df <- bind_rows(all_results) %>%
  mutate(date = as.Date(date)) %>%
  select(date, ecoregion, tmin, tmax, ppt, tdmean)

# write final historical climate data from PRISM
write.csv(climate_df, "Climate/outputs/hist_prism_800m_daily_MORA_03032026.csv", row.names = FALSE)

# Lets plots some of the results
OLYM_climate_df = read_csv("Climate/outputs/hist_prism_800m_daily_OLYM_02272026.csv")
MORA_climate_df = read_csv("Climate/outputs/hist_prism_800m_daily_MORA_03032026.csv")
NOCA_climate_df = read_csv("Climate/outputs/hist_prism_800m_daily_NOCA_03032026.csv")

# Aggregate to monthly tmax per ecoregion
OLYM_tmax_monthly_df <- OLYM_climate_df %>%
  filter(!is.na(tmax)) %>%
  mutate(
    year = year(date),
    month = month(date),
    ecoregion = factor(ecoregion),
    month_date = as.Date(paste(year, month, "15", sep = "-"))  # mid-month for x-axis
  ) %>%
  group_by(ecoregion, month_date) %>%
  summarize(tmax = mean(tmax, na.rm = TRUE), .groups = "drop")
# Aggregate to monthly tmax per ecoregion
OLYM_ppt_monthly_df <- OLYM_climate_df %>%
  filter(!is.na(tmax)) %>%
  mutate(
    year = year(date),
    month = month(date),
    ecoregion = factor(ecoregion),
    month_date = as.Date(paste(year, month, "15", sep = "-"))  # mid-month for x-axis
  ) %>%
  group_by(ecoregion, month_date) %>%
  summarize(ppt = mean(ppt, na.rm = TRUE), .groups = "drop")
# Aggregate to monthly tmax per ecoregion
MORA_tmax_monthly_df <- MORA_climate_df %>%
  filter(!is.na(tmax)) %>%
  mutate(
    year = year(date),
    month = month(date),
    ecoregion = factor(ecoregion),
    month_date = as.Date(paste(year, month, "15", sep = "-"))  # mid-month for x-axis
  ) %>%
  filter(year >= 1989) %>%
  group_by(ecoregion, month_date) %>%
  summarize(tmax = mean(tmax, na.rm = TRUE), .groups = "drop")
# Aggregate to monthly tmax per ecoregion
MORA_ppt_monthly_df <- MORA_climate_df %>%
  filter(!is.na(tmax)) %>%
  mutate(
    year = year(date),
    month = month(date),
    ecoregion = factor(ecoregion),
    month_date = as.Date(paste(year, month, "15", sep = "-"))  # mid-month for x-axis
  ) %>%
  filter(year >= 1989) %>%
  group_by(ecoregion, month_date) %>%
  summarize(ppt = mean(ppt, na.rm = TRUE), .groups = "drop")
# Aggregate to monthly tmax per ecoregion
NOCA_tmax_monthly_df <- NOCA_climate_df %>%
  filter(!is.na(tmax)) %>%
  mutate(
    year = year(date),
    month = month(date),
    ecoregion = factor(ecoregion),
    month_date = as.Date(paste(year, month, "15", sep = "-"))  # mid-month for x-axis
  ) %>%
  group_by(ecoregion, month_date) %>%
  summarize(tmax = mean(tmax, na.rm = TRUE), .groups = "drop")
# Aggregate to monthly tmax per ecoregion
NOCA_ppt_monthly_df <- NOCA_climate_df %>%
  filter(!is.na(tmax)) %>%
  mutate(
    year = year(date),
    month = month(date),
    ecoregion = factor(ecoregion),
    month_date = as.Date(paste(year, month, "15", sep = "-"))  # mid-month for x-axis
  ) %>%
  group_by(ecoregion, month_date) %>%
  summarize(ppt = mean(ppt, na.rm = TRUE), .groups = "drop")
# Aggregate to yearly tmax per ecoregion
MORA_tmax_annual_df <- MORA_climate_df %>%
  filter(!is.na(tmax)) %>%
  mutate(
    year = year(date),
    month = month(date),
    ecoregion = factor(ecoregion),
    month_date = as.Date(paste(year, month, "15", sep = "-"))  # mid-month for x-axis
  ) %>%
  filter(year >= 1989) %>%
  group_by(ecoregion, year) %>%
  summarize(tmax = mean(tmax, na.rm = TRUE), .groups = "drop")


# Plot with trend line
# OLYM maxtemp
ggplot(OLYM_tmax_monthly_df, aes(x = month_date, y = tmax, color = ecoregion)) +
  geom_line(alpha = 0.5, linewidth = 0.4) +
  geom_smooth(aes(group = 1), method = "loess", span = 0.2, se = FALSE, color = "black", linewidth = 1) +
  labs(
    title = "Monthly Mean Maximum Temperature by Ecoregion",
    subtitle = "PRISM 800m, OLYM (1981–2024)",
    x = "Year",
    y = "Tmax (°C)",
    color = "Ecoregion"
  ) +
  scale_x_date(date_breaks = "5 years", date_labels = "%Y") +
  scale_color_viridis_d() +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "right",
    panel.grid.minor = element_blank()
  )
# OLYM Pricip
ggplot(OLYM_ppt_monthly_df, aes(x = month_date, y = ppt, color = ecoregion)) +
  geom_line(alpha = 0.5, linewidth = 0.4) +
  geom_smooth(aes(group = 1), method = "loess", span = 0.2, se = FALSE, color = "black", linewidth = 1) +
  labs(
    title = "Monthly Precipitation by Ecoregion",
    subtitle = "PRISM 800m, OLYM (1981–2024)",
    x = "Year",
    y = "Precipitation (cm)",
    color = "Ecoregion"
  ) +
  scale_x_date(date_breaks = "5 years", date_labels = "%Y") +
  scale_color_viridis_d() +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "right",
    panel.grid.minor = element_blank()
  )
# MORA maxtemp
ggplot(MORA_tmax_monthly_df, aes(x = month_date, y = tmax, color = ecoregion)) +
  geom_line(alpha = 0.5, linewidth = 0.4) +
  geom_smooth(aes(group = 1), method = "loess", span = 0.2, se = FALSE, color = "black", linewidth = 1) +
  labs(
    title = "Monthly Mean Maximum Temperature by Ecoregion",
    subtitle = "PRISM 800m, MORA (1981–2024)",
    x = "Year",
    y = "Tmax (°C)",
    color = "Ecoregion"
  ) +
  scale_x_date(date_breaks = "5 years", date_labels = "%Y") +
  scale_color_viridis_d() +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "right",
    panel.grid.minor = element_blank()
  )
# MORA Pricip
ggplot(MORA_ppt_monthly_df, aes(x = month_date, y = ppt, color = ecoregion)) +
  geom_line(alpha = 0.5, linewidth = 0.4) +
  geom_smooth(aes(group = 1), method = "loess", span = 0.2, se = FALSE, color = "black", linewidth = 1) +
  labs(
    title = "Monthly Precipitation by Ecoregion",
    subtitle = "PRISM 800m, MORA (1981–2024)",
    x = "Year",
    y = "Precipitation (cm)",
    color = "Ecoregion"
  ) +
  scale_x_date(date_breaks = "5 years", date_labels = "%Y") +
  scale_color_viridis_d() +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "right",
    panel.grid.minor = element_blank()
  )
# NOCA maxtemp
ggplot(NOCA_tmax_monthly_df, aes(x = month_date, y = tmax, color = ecoregion)) +
  geom_line(alpha = 0.5, linewidth = 0.4) +
  geom_smooth(aes(group = 1), method = "loess", span = 0.2, se = FALSE, color = "black", linewidth = 1) +
  labs(
    title = "Monthly Mean Maximum Temperature by Ecoregion",
    subtitle = "PRISM 800m, NOCA (1981–2024)",
    x = "Year",
    y = "Tmax (°C)",
    color = "Ecoregion"
  ) +
  scale_x_date(date_breaks = "5 years", date_labels = "%Y") +
  scale_color_viridis_d() +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "right",
    panel.grid.minor = element_blank()
  )
# NOCA Pricip
ggplot(NOCA_ppt_monthly_df, aes(x = month_date, y = ppt, color = ecoregion)) +
  geom_line(alpha = 0.5, linewidth = 0.4) +
  geom_smooth(aes(group = 1), method = "loess", span = 0.2, se = FALSE, color = "black", linewidth = 1) +
  labs(
    title = "Monthly Precipitation by Ecoregion",
    subtitle = "PRISM 800m, NOCA (1981–2024)",
    x = "Year",
    y = "Precipitation (cm)",
    color = "Ecoregion"
  ) +
  scale_x_date(date_breaks = "5 years", date_labels = "%Y") +
  scale_color_viridis_d() +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "right",
    panel.grid.minor = element_blank()
  )
# NOCA tmax annual
ggplot(MORA_tmax_annual_df, aes(x = as.Date(year), y = tmax, color = ecoregion)) +
  geom_line(alpha = 0.5, linewidth = 0.4) +
  geom_smooth(aes(group = 1), method = "loess", span = 0.2, se = FALSE, color = "black", linewidth = 1) +
  labs(
    title = "Annual Tmax by Ecoregion",
    subtitle = "PRISM 800m, NOCA (1981–2024)",
    x = "Year",
    y = "Average Maximum Temp (C)",
    color = "Ecoregion"
  ) +
  scale_x_date(date_breaks = "5 years", date_labels = "%Y") +
  scale_color_viridis_d() +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "right",
    panel.grid.minor = element_blank()
  )

# They look good, but there giant gaps in years before 1988. Lets filter for 1988 - 2024 and also reformat for the correct LANDIS climate structure.
# OLYM
OLYM_landis_climate_df <- OLYM_climate_df %>%
  filter(year(date) >= 1988) %>%
  mutate(
    Year = year(date),
    Month = month(date),
    Day = day(date)
  ) %>%
  pivot_longer(
    cols = c(tmin, tmax, ppt, tdmean),
    names_to = "Variable",
    values_to = "Value"
  ) %>%
  select(Year, Month, Day, Variable, ecoregion, Value) %>%
  pivot_wider(
    names_from = ecoregion,
    values_from = Value,
    names_sort = TRUE
  ) %>%
  arrange(Year, Month, Day, Variable) %>%
  mutate(Variable = case_when(
    Variable == "tdmean" ~ "dewpoint",
    .default = Variable
  ))
# MORA
MORA_landis_climate_df <- MORA_climate_df %>%
  filter(year(date) >= 1988) %>%
  mutate(
    Year = year(date),
    Month = month(date),
    Day = day(date)
  ) %>%
  pivot_longer(
    cols = c(tmin, tmax, ppt, tdmean),
    names_to = "Variable",
    values_to = "Value"
  ) %>%
  select(Year, Month, Day, Variable, ecoregion, Value) %>%
  pivot_wider(
    names_from = ecoregion,
    values_from = Value,
    names_sort = TRUE
  ) %>%
  arrange(Year, Month, Day, Variable) %>%
  mutate(Variable = case_when(
    Variable == "tdmean" ~ "dewpoint",
    .default = Variable
  ))
# NOCA
NOCA_landis_climate_df <- NOCA_climate_df %>%
  filter(year(date) >= 1988) %>%
  mutate(
    Year = year(date),
    Month = month(date),
    Day = day(date)
  ) %>%
  pivot_longer(
    cols = c(tmin, tmax, ppt, tdmean),
    names_to = "Variable",
    values_to = "Value"
  ) %>%
  select(Year, Month, Day, Variable, ecoregion, Value) %>%
  pivot_wider(
    names_from = ecoregion,
    values_from = Value,
    names_sort = TRUE
  ) %>%
  arrange(Year, Month, Day, Variable) %>%
  mutate(Variable = case_when(
    Variable == "tdmean" ~ "dewpoint",
    .default = Variable
  ))

# Export to CSV
write_csv(OLYM_landis_climate_df, "Climate/outputs/hist_prism_800m_daily_1988-2024_OLYM_03032026.csv")
write_csv(MORA_landis_climate_df, "Climate/outputs/hist_prism_800m_daily_1988-2024_MORA_03032026.csv")
write_csv(NOCA_landis_climate_df, "Climate/outputs/hist_prism_800m_daily_1988-2024_NOCA_03032026.csv")
```

After running this through LANDIS it looks like there might be some missing dates in the dataset. Lets take a look and see how many and if we can easily fill them.
```{r}
# Load required libraries
library(dplyr)
library(readr)
library(lubridate)
setwd("/Users/jlamping/Dropbox (University of Oregon)/Lamping/NPS_postdoc")

# Load original climate dataset. Loop through and do this for the different landscapes
climate_df <- read_csv("Spatial/Climate/outputs/hist_prism_800m_daily_1988-2024_NOCA_03032026.csv") %>%
  mutate(Date = make_date(Year, Month, Day),
         Source = "original") %>% # flag original rows
  filter(Year >= 1989) %>%
  select(-eco0)

# Create complete daily date sequence
expected_dates <- tibble(Date = seq.Date(as.Date("1989-01-01"), as.Date("2024-12-31"), by = "day"))

# Extract all unique variables
variables <- unique(climate_df$Variable)

# Build full expected (Date × Variable) matrix
expected_matrix <- expand.grid(Date = expected_dates$Date,
                               Variable = variables,
                               stringsAsFactors = FALSE) %>%
  as_tibble() %>%
  mutate(Month = month(Date),
         Day   = day(Date))

# Get actual (Date, Variable) pairs present
actual_keys <- climate_df %>%
  select(Date, Variable) %>%
  distinct()

# Identify missing (Date, Variable) pairs
missing_matrix <- anti_join(expected_matrix, actual_keys, by = c("Date", "Variable"))

# Add Month & Day to main dataset for sampling
climate_df <- climate_df %>%
  mutate(Month = month(Date),
         Day   = day(Date))

# Fill missing by randomly sampling from other years with same Month/Day/Variable
# filled_rows <- missing_matrix %>%
#   select(Date, Variable) %>%
#   rowwise() %>%
#   mutate(Fill_Row = list({
#     pool <- climate_df %>%
#       filter(Variable == Variable,
#              Month == month(Date),
#              Day == day(Date))
# 
#     #if (nrow(pool) == 0) return(NULL)
# 
#     sample_row <- sample_n(pool, 1) %>%
#       mutate(Year = year(Date),
#              Month = month(Date),
#              Day = day(Date),
#              Date = Date,
#              Source = "filled")
#     sample_row
#   })) %>%
#   tidyr::unnest(Fill_Row, names_sep = "_")

filled_rows <- missing_matrix %>%
  select(Date, Variable) %>%
  rowwise() %>%
  mutate(Fill_Row = list({

    current_var  <- Variable
    current_date <- Date

    pool <- climate_df %>%
      filter(
        Variable == current_var,
        Month == month(current_date),
        Day   == day(current_date)
      )

    sample_row <- sample_n(pool, 1) %>%
      mutate(
        Year  = year(current_date),
        Month = month(current_date),
        Day   = day(current_date),
        Date  = current_date,
        Source = "filled"
      )

    sample_row
  })) %>%
  tidyr::unnest(Fill_Row, names_sep = "_")

# Reformat the unnested filled rows to match climate_df column names
# filled_rows_clean <- filled_rows %>%
#   transmute(
#     Year    = year(Date),
#     Month   = month(Date),
#     Day     = day(Date),
#     Variable = Variable,
#     Source   = Fill_Row_Source,
#     Date     = Date,
#     across(starts_with("Fill_Row_eco"), .names = "{str_replace(.col, 'Fill_Row_', '')}")
#   )
filled_rows_clean <- filled_rows %>%
  transmute(
    Year    = year(Date),
    Month   = month(Date),
    Day     = day(Date),
    Variable = Variable,
    Source   = Fill_Row_Source,
    Date     = Date,
    across(starts_with("Fill_Row_eco"), 
           .names = "{str_replace(.col, 'Fill_Row_', '')}")
  )

# Now combine with the original dataset
climate_complete <- bind_rows(climate_df, filled_rows_clean) %>%
  arrange(Date) %>% select(-c(Date, Source))

# # Optional: check that it's complete now
# expected_final <- expand.grid(Date = expected_dates$Date,
#                               Variable = variables,
#                               stringsAsFactors = FALSE)
# 
# missing_after <- anti_join(expected_final,
#                            climate_complete %>% select(Date, Variable),
#                            by = c("Date", "Variable"))
# 
# stopifnot(nrow(missing_after) == 0)

# Convert only ppt values from mm to cm
# Get the eco columns (assuming they are named exactly "eco1" to "eco22")
eco_cols <- paste0("eco", 1:22)
# Convert only ppt rows for those columns
climate_complete[climate_complete$Variable == "ppt", eco_cols] <- 
  climate_complete[climate_complete$Variable == "ppt", eco_cols] / 10
# # No missing dates for MORA, just fix ppt and write
# climate_df[climate_df$Variable == "ppt", eco_cols] <- 
#   climate_df[climate_df$Variable == "ppt", eco_cols] / 10
# LOOKS GOOOOOOOOOD!!!
write_csv(climate_complete, "Spatial/Climate/outputs/hist_prism_800m_daily_1989-2024_MORA_03032026_filled.csv")

# # write without dewpoint for initial testing
# write_csv(climate_complete %>% filter(Variable != "tdmean"),
#           "LANDIS/initial_landscapes/OLYM/extensions/hist_prism_800m_daily_1988-2024_OLYM_NoDew.csv")
```

# Future climate
```{r}
################################################################################
#  NEX‑GDDP‑CMIP6 ▸ Washington State subset ▸ R version
#  Models  : CanESM5, HadGEM3‑GC31‑LL, CNRM‑CM6‑1, GFDL‑ESM4, MRI‑ESM2‑0
#  Scenarios: ssp126 (SSP1‑2.6), ssp245 (SSP2‑4.5), ssp370 (SSP3‑7.0)
#  Variables: tas tasmax tasmin pr rsds rlds hurs huss sfcWind
#  Period  : 2015‑01‑01 to 2100‑12‑31
#  Output  : one cropped NetCDF per model‑scenario‑variable
################################################################################

if (!require("pacman")) install.packages("pacman")
pacman::p_load("aws.s3", "terra", "stringr", "dplyr", "purrr", 'progress', 'glue', 'ncdf4', "furrr", "future", "prism", "exactextractr", "sf", "data.table", "future.apply") # these are the libraries we are using
#pak::pkg_install("mt-climate-office/cmip6@*release")
# set working directory where you want the data downloaded
setwd("/Volumes/BigStore/PNW_Parks/Spatial/Climate/")
setwd("/Users/jlamping/Dropbox (University of Oregon)/Lamping/NPS_postdoc/Spatial/Climate/")
setwd("/Users/jameslamping/Dropbox (University of Oregon)/Lamping/NPS_postdoc/Spatial/Climate/")
options(scipen=999) # no sci notation
```

Script Description: 
  This section of code queries the NASA NEX-GDDP-CMIP6 archive hosted on AWS, 
  identifies valid daily NetCDF files for selected GCMs, scenarios, and variables, 
  and downloads them in parallel using {furrr}. 

  For each file, it:
    1. Crops the raster to a bounding box around Washington State,
    2. Subsets to daily data from 2015–2100,
    3. Rotates longitudes from 0–360 to -180–180,
    4. Writes the cropped and cleaned NetCDF to disk using LANDIS-compatible naming.

  Output:
    One NetCDF file per model × scenario × variable covering 2015–2100 and 
    spatially cropped to the study region.

  Dependencies:
    - aws.s3, terra, glue, furrr, future, dplyr, stringr
    - AWS public S3 access (no credentials required)

```{r}
Sys.setenv("AWS_DEFAULT_REGION" = "us-west-2")
bucket <- "nex-gddp-cmip6"

# ── User-specified inputs ─────────────────────────────────────────────────────
models <- c("CNRM-CM6-1", "GFDL-ESM4", "MRI-ESM2-0", "CanESM5", "HadGEM3-GC31-LL",
            "ACCESS-CM2", "ACCESS-ESM1-5", "CMCC-ESM2", "CNRM-ESM2-1", "EC-Earth3",
            "EC-Earth3-Veg-LR", "FGOALS-g3", "GISS-E2-1-G", "INM-CM4-8", "INM-CM5-0",
            "KACE-1-0-G", "MIROC-ES2L", "MPI-ESM1-2HR", "MPI-ESM1-2-LR", "NorESM2-LM",
            "NorESM2-MM", "TaiESM1")
scenarios  <- c("ssp370", "ssp585")
# variant_map <- c(
#   "CNRM-CM6-1" = "r1i1p1f2",
#   "GFDL-ESM4" = "r1i1p1f1",
#   "MRI-ESM2-0" = "r1i1p1f1",
#   "CanESM5" = "r1i1p1f1",
#   "HadGEM3-GC31-LL" = "r1i1p1f3"
# )

#vars_keep <- c("tas","tasmax","tasmin","pr","rsds","rlds","hurs","huss","sfcWind")
vars_keep <- c("tasmax","tasmin", "pr", "huss","sfcWind")
bbox <- ext(-125, -116, 45, 49)
bbox <- vect(bbox, crs = "epsg:4326")
bbox_native <- project(bbox, crs("OGC:CRS84"))
out_dir   <- "./nex_gddp_wa_3"
#dir.create(out_dir, showWarnings = FALSE)

# ── Step 1: Collect valid file paths from AWS ─────────────────────────────────
get_valid_keys_df <- function(models, scenarios, vars_keep) {
  # Build prefixes for all model-scenario-variant combinations
  prefixes <- expand.grid(models = models, scenarios = scenarios) %>%
    #mutate(variant = variant_map[models]) %>%
    mutate(prefix = glue("NEX-GDDP-CMIP6/{models}/{scenarios}/")) %>%
    #mutate(prefix = glue("NEX-GDDP-CMIP6/{models}/{scenarios}/{variant}/")) %>%
    pull(prefix)
  
  # Pull all file keys for the model-scenario-variant combos
  all_files <- bind_rows(
    lapply(prefixes, function(prefix) {
      rbindlist(
        get_bucket(
              bucket = bucket,
              prefix = prefix,
              max = Inf,
              use_https = TRUE,
              region = "us-west-2",
              url_style = "path"
            ),
        use.names = TRUE, fill = TRUE) #%>% select(Key)
    })
  )
  
  # Add columns for filtering
  all_files <- all_files %>%
    mutate(model = str_extract(Key, paste(models, collapse = "|")),
           scenario = str_extract(Key, paste(scenarios, collapse = "|")),
           variant = str_split(Key, "/", simplify = TRUE)[, 4],
           variable = str_extract(Key, paste(vars_keep, collapse = "|")),
           version = str_extract(Key, "v[0-9]+\\.[0-9]+")) 
  
  return(all_files)
}

all_files_df <- get_valid_keys_df(models, scenarios, vars_keep)

# Filter to just v2.0 files
valid_files_df <- all_files_df %>%
  filter(version == "v2.0") %>% filter(!is.na(scenario)) %>% filter(!is.na(variable))
  #select(Key)

# Create summary table to check that we have all the scenarios and variables we want for each model
print(
  valid_files_df %>%
  distinct(model, scenario, variable) %>%
  group_by(model) %>%
  summarise(
    scenarios = toString(sort(unique(scenario))),
    variables = toString(sort(unique(variable))),
    .groups = "drop")
  )

# Extract just the "key" column as a character vector for downloading
all_keys <- valid_files_df$Key

length(all_keys)  # sanity check

# Funtion to download and process all identified AWS paths using future
process_one_file <- function(key) {
  # Ensure bucket is globally available
  bucket <- "nex-gddp-cmip6"
  out_dir <- out_dir
  bbox <- vect(ext(-125, -116, 45, 49), crs = "epsg:4326")
  bbox_native <- project(bbox, crs("OGC:CRS84"))

  # Extract components from the key
  #key = all_keys[1]
  filename <- basename(key)
  parts <- unlist(str_split(filename, "_"))
  var <- parts[1]
  model <- parts[3]
  scen <- parts[4]

  # Set output file path
  out_file <- file.path(out_dir, glue("{model}_{scen}_{var}_WA_{parts[7]}"))

  # Skip if already processed
  if (file.exists(out_file)) {
    message("✅ Already exists: ", out_file)
    return(NULL)
  }

  # Process the NetCDF file
  tmp <- tempfile(fileext = ".nc")
  tryCatch({
    save_object(key, bucket = bucket, file = tmp,
                use_https = TRUE, region = "us-west-2")
    r <- rast(tmp)
    time_index <- time(r)
    keep_times <- which(time_index >= as.Date("2015-01-01") & time_index <= as.Date("2100-12-31"))
    r <- subset(r, keep_times)
    r <- terra::rotate(r)  # Fix longitude
    r <- crop(r, bbox_native, snap = "out")
    writeCDF(r, paste0(out_file, ".nc"),
             overwrite = TRUE,
             varname = var, xname = "lon", yname = "lat",
             varunit = terra::units(r)[1])
    unlink(tmp)
    unlink(r)
    #message("✅ Wrote: ", out_file)
  }, error = function(e) {
    message("❌ Failed: ", key)
  })
  r
}

# Apply to all valid keys on single core
#walk(all_keys[1:2], process_one_file)

# Set up parallel plan
plan(multisession, workers = 18)  # adjust to your CPU

# Run in parallel with progress bar
future_walk(all_keys, process_one_file, .progress = TRUE)

```

# Download check

Lets run a quick check to see if we have what we need for all the models we wanted.
```{r}
library(tidyverse)
library(fs)
library(stringr)
library(ggplot2)
setwd("/Volumes/BigStore/PNW_Parks/Spatial/Climate")

# Adjust this path to where your downloaded NetCDF files are stored
output_dir <- "nex_gddp_wa_2"

# List all downloaded files
files <- dir_ls(output_dir, glob = "*.nc")

# Extract metadata: model, scenario, variable, year
file_info <- tibble(file = basename(files)) %>%
  mutate(
    model     = str_extract(file, "^[^_]+"),
    scenario  = str_extract(file, "(ssp\\d{3})"),
    variable  = str_extract(file, "(tas|tasmax|tasmin|pr|rsds|rlds|hurs|huss|sfcWind)"),
    year      = str_extract(file, "_20\\d{2}\\.nc$") %>% str_remove_all("_|\\.nc") %>% as.integer(),
    group     = paste(model, scenario, sep = "_")
  )

# Create presence matrix (1 = file exists for that year)
coverage_df <- file_info %>%
  distinct(group, year) %>%
  mutate(present = 1)

# Plot: presence matrix across years
ggplot(coverage_df, aes(x = year, y = fct_rev(group), fill = factor(present))) +
  geom_tile(color = "gray80") +
  scale_fill_manual(values = c("1" = "forestgreen"), labels = c("Data Present")) +
  labs(title = "Downloaded NEX-GDDP Climate Data Coverage by Model–Scenario",
       x = "Year", y = "Model–Scenario", fill = "") +
  theme_minimal(base_size = 14)
```

# This script processes daily NASA NEX-GDDP-CMIP6 NetCDF climate files for a given national park,
# extracting spatial averages by ecoregion zone, converting them to LANDIS-II-compatible units and format,
# and writing one CSV file per model–scenario group. The code performs spatial reprojection, resampling,
# and zonal mean calculations in parallel using furrr and terra, and includes unit conversions 
# (e.g., K to °C, mm/s to cm/day). Output files are structured by date and ecoregion for direct use 
# in LANDIS-II climate libraries.

```{r}
library(terra)
library(dplyr)
library(stringr)
library(lubridate)
library(glue)
library(furrr)
library(future)
library(progressr)
library(tidyr)
library(tidyverse)

setwd("/Volumes/BigStore/PNW_Parks/Spatial/Climate/")
setwd("/Users/jlamping/Dropbox (University of Oregon)/Lamping/NPS_postdoc/Spatial/Climate/")
setwd("/Users/jameslamping/Dropbox (University of Oregon)/Lamping/NPS_postdoc/Spatial/Climate/")
options(scipen=999) # no sci notation

# --- Paths ---
park = "NOCA"
nc_dir <- "/Volumes/BigStore/PNW_Parks/Spatial/Climate/nex_gddp_wa_2"  # Folder with processed NetCDFs
eco_raster_path <- glue("../../LANDIS/initial_landscapes/{park}/input_maps/{park}_ClimateRegions_09122025.tif")
output_dir <- glue("outputs/Future_climate_processed/{park}")
#dir.create(output_dir, showWarnings = FALSE)
eco_r <- rast(eco_raster_path)  # load once instead of every day

# --- List all NetCDFs ---
nc_files <- list.files(nc_dir, pattern = "\\.nc$", full.names = TRUE)

# --- Group files by model–scenario ---
file_info <- tibble(file = nc_files) %>%
  mutate(
    name = basename(file),
    model = str_split_i(name, "_", 1),
    scenario = str_split_i(name, "_", 2),
    variable = str_split_i(name, "_", 3),
    year = str_split_i(name, "_", 5),
    group = glue("{model}_{scenario}")
    ) %>%
  mutate(
    year = as.integer(str_remove(year, ".nc"))) %>%
  filter(scenario != "ssp245") %>% # not analyzing this scenario
  filter(model != "CMCC-ESM2") %>%
  filter(model != "MIROC-ES2L")

# --- Loop over each group (model–variant–scenario) ---
# --- Process each model-scenario group ---
# ── Parallel plan ─────────────────────────────────────────────────────
#plan(multisession, workers = 10)

groups <- unique(file_info$group)
groups = groups[-c(20, 25:26)]
plan(multisession, workers = 8) #parallel::detectCores() - 10
# --- Loop over each group (model + scenario) ---
for (g in groups) {
  #g = "ACCESS-CM2_ssp370"
  message("Processing group: ", g)
  sub_files <- file_info %>% filter(group == g)

  with_progress({
    p <- progressor(along = 1:nrow(sub_files))
    
    # process variable and years in parallel
    group_df <- future_map_dfr(seq_len(nrow(sub_files)), function(i) {
      #i = 1
      park = "NOCA"
      f = sub_files[i, ]$file
      model = sub_files[i, ]$model
      scenario = sub_files[i, ]$scenario
      variable = sub_files[i, ]$variable
      # Read stack and extract time info
      eco_r <- rast(glue("../../LANDIS/initial_landscapes/{park}/input_maps/{park}_ClimateRegions_09122025.tif"))
      names(eco_r) = "ecoregion"
      r <- rast(f) #
      dates <- as.Date(time(r))
      names(r) <- dates
      r = project(r, crs(eco_r))
      r = resample(r, eco_r, method = "bilinear", threads = FALSE)
      z <- as.data.frame(zonal(r, eco_r, fun = "mean", na.rm = TRUE))
      rm(r)
      terra::tmpFiles(remove = TRUE)
      # Convert to long format
      z_long <- z %>%
        pivot_longer(
          cols = -ecoregion,
          names_to = "date",
          values_to = "value"
          ) %>%
        mutate(
          date = as.Date(date),
          variable = variable,
          model = model,
          scenario = scenario
          )
      p()
      return(z_long)}
      ) # .options = furrr_options(globals = list(park = park, eco_r = eco_r, sub_files = sub_files), seed = TRUE)

        ################################################################################
    # Unit conversions: NASA NEX‑GDDP‑CMIP6  →  LANDIS‑II Climate Library
    #
    # Variable mapping + unit changes
    #
    # NASA name   → LANDIS name   | NASA units              → LANDIS units   | Conversion
    # ------------------------------------------------------------------------------
    # pr          → ppt           | kg m-2 s-1 (mm/s)       → cm/day         | value * 86400 / 10
    # tasmax      → maxtemp       | Kelvin (K)              → °C             | value - 273.15
    # tasmin      → mintemp       | Kelvin (K)              → °C             | value - 273.15
    # huss        → SH            | kg/kg                   → g/kg           | value * 1000
    # sfcWind     → windSpeed     | m/s                     → km/h           | value * 3.6
    #
    # Notes:
    # • 1 kg m-2 water = 1 mm precipitation
    # • 86400 sec/day converts rate → daily accumulation
    # • ÷10 converts mm → cm
    # • LANDIS expects:
    #     ppt       = cm/day
    #     temps     = °C
    #     SH        = g/kg
    #     windSpeed = km/h
    ################################################################################
    final_df = group_df %>%
        mutate(
          Year = year(date),
          Month = month(date),
          Day = day(date)
        ) %>%
        select(Year, Month, Day, variable, ecoregion, value) %>%
        pivot_wider(
          names_from = ecoregion,
          values_from = value,
          names_prefix = "eco"
        ) %>%
      # change names and units to what LANDIS needs
      rename(Variable = variable) %>%
      # variable names
      mutate(Variable = case_when(
          Variable == "pr" ~ "ppt",
          Variable == "tasmax" ~ "maxtemp",
          Variable == "tasmin" ~ "mintemp",
          Variable == "huss" ~ "SH", # specific humidity
          Variable == "sfcWind" ~ "windSpeed",
          .default = Variable)) %>%
      # units
      mutate(across(starts_with("eco"), ~ case_when(
        Variable == "ppt"       ~ .x * 86400 / 10,
        Variable == "maxtemp"   ~ .x - 273.15,
        Variable == "mintemp"   ~ .x - 273.15,
        Variable == "SH"        ~ .x * 1000,
        Variable == "windSpeed" ~ .x * 3.6,
        TRUE ~ .x)))
    }) 
   
  
  # Save 
  write_csv(final_df, glue("{output_dir}/{g}_{park}.csv"))
  message("Saved: ", glue("{output_dir}/{g}_{park}.csv"))
    
  # Clean memory after each group
  gc(verbose = FALSE)
}

```
# Model comparison

# This workflow ingests LANDIS-ready climate CSV outputs (daily values by ecoregion),
# reshapes them to long format, computes spatial means across ecoregions, and
# aggregates to annual metrics. Annual temperature is calculated as the mean of daily
# values and precipitation as the annual sum (converted to cm). The script then:
#   1) plots annual time-series comparisons across all model–scenario combinations, and
#   2) computes early vs late period deltas (ΔT and ΔP%) to support a four-corners
#      climate envelope selection for representative GCM/SSP scenarios.

```{r}
library(tidyverse)
library(lubridate)
library(fs)
library(ggrepel)
library(patchwork)

# Path to folder with all OLYM CSVs
climate_dir <- "outputs/Future_climate_processed/NOCA"

# Load all climate files
files <- dir_ls(climate_dir, regexp = "_NOCA\\.csv$")

# Combine all into one dataframe
climate_df <- map_dfr(files, function(f) {
  df <- read_csv(f, show_col_types = FALSE)
  meta <- str_match(basename(f), "^(.*?)_(.*?)_NOCA\\.csv$")
  df$model <- meta[,2]
  df$scenario <- meta[,3]
  df
}) %>% select(-eco0)

# Average across ecoregions per day
climate_daily <- climate_df %>%
  pivot_longer(cols = starts_with("eco"), names_to = "ecozone", values_to = "value") %>%
  #   mutate(variable = case_when(
  #   variable == "pr" ~ "ppt",
  #   variable == "tasmax" ~ "tmax",
  #   variable == "tasmin"~ "tmin"
  # )) %>%
  group_by(model, scenario, Variable, Year, Month, Day) %>%
  summarise(value = mean(value, na.rm = TRUE), .groups = "drop")

# Compute annual mean per model/scenario/variable
climate_annual <- climate_daily %>%
  group_by(model, scenario, Variable, Year) %>%
  summarise(annual_value = mean(value, na.rm = TRUE), .groups = "drop")

# Compute annual values: sum for pr, mean for other variables
# cm/yr = pr * 31,5557,600 seconds * 0.1 = pr * 3,155,760
climate_annual_pr <- climate_daily %>%
  # mutate(value = case_when(
  #   variable == "ppt" ~ value * 86400 / 10,
  #   .default = value)) %>%
  group_by(model, scenario, Variable, Year) %>%
  summarise(
    annual_value = if_else(
      Variable == "ppt",
      sum(value, na.rm = TRUE),
      mean(value, na.rm = TRUE)
    )[1],
    , .groups = "drop") #%>%
 

# ─── PLOT: tasmax comparison ────────────────────────────────────────────────
climate_annual_pr %>%
  filter(Variable == "maxtemp") %>%
  ggplot(aes(x = Year, y = annual_value, color = interaction(model, scenario))) +
  geom_line(linewidth = 1) +
  labs(title = "Annual Average tasmax (°K)",
       y = "Temperature (°K)", x = "Year", color = "Model–Scenario") +
  theme_minimal()

# ─── PLOT: precipitation comparison ─────────────────────────────────────────
climate_annual_pr %>%
  filter(Variable == "ppt") %>%
  ggplot(aes(x = Year, y = annual_value, color = interaction(model, scenario))) +
  geom_line(linewidth = 1) +
  labs(title = "Total Annual Precipitation",
       y = "Precipitation (cm)", x = "Year", color = "Model–Scenario") +
  theme_minimal()

# ─── FOUR CORNERS SELECTION ─────────────────────────────────────────────────
# ---- SETTINGS (EDIT THESE) ----
begin_years <- 2015:2044
end_years   <- 2071:2100
vars <- c("maxtemp", "ppt")

# ---- 1) PERIOD MEANS (BEGIN vs END) ----
period_means <- climate_annual_pr %>%
  filter(Variable %in% vars, Year >= 2015, Year <= 2100) %>%
  mutate(period = case_when(
    Year %in% begin_years ~ "begin",
    Year %in% end_years   ~ "end",
    TRUE ~ NA_character_
  )) %>%
  filter(!is.na(period)) %>%
  group_by(model, scenario, Variable, period) %>%
  summarise(period_mean = mean(annual_value, na.rm = TRUE), .groups = "drop")

# ---- 2) WIDE + DELTAS ----
deltas <- period_means %>%
  pivot_wider(names_from = period, values_from = period_mean) %>%
  filter(!is.na(begin), !is.na(end)) %>%
  mutate(
    # Temperature: if tmax is in Kelvin, ΔK == Δ°C (so this is Δ°C)
    delta_abs = end - begin,
    # Precip: percent change relative to begin
    delta_pct = 100 * (end - begin) / begin
  )

corners_df <- deltas %>%
  select(model, scenario, Variable, delta_abs, delta_pct) %>%
  pivot_wider(
    names_from = Variable,
    values_from = c(delta_abs, delta_pct),
    names_sep = "_"
  ) %>%
  transmute(
    model,
    scenario,
    delta_T_C   = delta_abs_maxtemp,
    delta_P_pct = delta_pct_ppt
  ) %>%
  filter(!is.na(delta_T_C), !is.na(delta_P_pct))

# ---- 3) PLOT (ALL MODELS + ALL SSPS) ----
# plot for each park
NOCA_corner_plt = ggplot(corners_df, aes(x = delta_P_pct, y = delta_T_C, color = scenario)) +
  geom_hline(yintercept = 0, linewidth = 0.4, color = "grey40") +
  geom_vline(xintercept = 0, linewidth = 0.4, color = "grey40") +
  geom_point(size = 3, alpha = 0.9) +
  ggrepel::geom_text_repel(
    aes(label = model),
    size = 3,
    max.overlaps = Inf,
    box.padding = 0.25,
    point.padding = 0.15
  ) +
  labs(
    title = paste0(
      "NOCA Projected Precip & Temp Change (All SSPs)\n",
      min(begin_years), "–", max(begin_years),
      " to ",
      min(end_years), "–", max(end_years)
    ),
    x = expression(Delta~"P (%)"),
    y = expression(Delta~"T ("*degree*"C)"),
    color = "Scenario"
  ) +
  theme_minimal()

############################ Model Decision
# Warm/Dry = ACCESS-CM2 ssp585
# Warm/Wet = TaiESM1 ssp585
# Cool/Dry = GISS-E2-1-G ssp370
# Cool/Wet = INM-CM5-0 ssp370

# stack plots
(OLYM_corner_plt/MORA_corner_plt/NOCA_corner_plt) +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")

## adding historical data from PRISM
# Load historical PRISM data
prism_path <- "../../LANDIS/initial_landscapes/MORA/extensions/hist_prism_800m_daily_1988-2024_MORA_filled.csv"
hist_df <- read_csv(prism_path, show_col_types = FALSE)

# Add model and scenario columns
hist_df <- hist_df %>%
  mutate(model = "PRISM", scenario = "historical") %>%
  rename(variable = Variable)

# Combine historical + future
climate_df_full <- bind_rows(climate_df, hist_df) %>%
  mutate(variable = case_when(
    variable == "pr" ~ "ppt",
    variable == "tasmax" ~ "tmax",
    variable == "tasmin"~ "tmin",
    .default = variable
  ))

# Average across ecoregions per day
climate_daily <- climate_df_full %>%
  pivot_longer(cols = starts_with("eco"), names_to = "ecozone", values_to = "value") %>%
  group_by(model, scenario, variable, Year, Month, Day) %>%
  summarise(value = mean(value, na.rm = TRUE), .groups = "drop")

# Annual mean tasmax etc.
climate_annual <- climate_daily %>%
  group_by(model, scenario, variable, Year) %>%
  summarise(mean_value = mean(value, na.rm = TRUE), .groups = "drop") %>%
  mutate(mean_value = case_when(
    variable == "tmax" & model != "PRISM" ~ mean_value - 273.15,  # Convert kg/m²/s → cm/day
    .default = mean_value # PRISM already in mm/day → cm/day
  ))

# Annual precipitation totals (converted to cm)
climate_annual_pr <- climate_daily %>%
  mutate(value = case_when(
    variable == "ppt" & model != "PRISM" ~ value * 86400 / 10,  # Convert kg/m²/s → cm/day
    TRUE ~ value  # PRISM already in mm/day → cm/day
  )) %>%
  group_by(model, scenario, variable, Year) %>%
  summarise(annual_value = sum(value, na.rm = TRUE), .groups = "drop") %>%
  filter(model != "HadGEM3-GC31-LL")


# TASMAX PLOT
climate_annual %>%
  filter(variable == "tmax") %>%
  ggplot(aes(x = Year, y = mean_value, color = interaction(model, scenario))) +
  geom_line(linewidth = 1) +
  labs(title = "Annual Average tasmax (°C)",
       y = "Temperature (°C)", x = "Year", color = "Model–Scenario") +
  theme_minimal()

# PRECIPITATION PLOT
climate_annual_pr %>%
  filter(variable == "ppt") %>%
  ggplot(aes(x = Year, y = annual_value, color = interaction(model, scenario))) +
  geom_line(linewidth = 1) +
  labs(title = "Total Annual Precipitation (Historical + Future)",
       y = "Precipitation (cm)", x = "Year", color = "Model–Scenario") +
  theme_minimal()
```

# Climate Stream Figure - MaxTemp
```{r}
################################################################################
# plot_NOCA_maxtemp_all_models.R
#
# Plots annual mean daily maximum temperature for:
#   - All future climate model CSVs in the NOCA directory
#   - Historic PRISM baseline
#
# Future CSVs: Variable == "maxtemp", spatial columns eco0:eco22
# Historic CSV: Variable == "tmax",   spatial columns eco0:eco22
#
# Four "decision" models are highlighted; all others shown in grey.
################################################################################
 
library(tidyverse)
library(scales)
 
# ── Paths ─────────────────────────────────────────────────────────────────────
future_dir  <- "/Users/jlamping/University of Oregon Dropbox/James Lamping/Lamping/NPS_postdoc/Spatial/Climate/outputs/Future_climate_processed/NOCA"
historic_fp <- "/Users/jlamping/University of Oregon Dropbox/James Lamping/Lamping/NPS_postdoc/Spatial/Climate/outputs/hist_prism_800m_daily_1988-2024_NOCA_03032026.csv"
 
# ── Helper: read one CSV and return annual mean tmax ──────────────────────────
# Averages across all eco columns, then across all days within each year.
read_annual_tmax <- function(filepath, var_label, tmax_varname) {
  message("Reading: ", basename(filepath))
  
  df <- read_csv(filepath, show_col_types = FALSE)
  
  # Keep only rows matching the tmax variable
  df <- df %>% filter(Variable == tmax_varname)
  
  # Identify eco columns (all numeric columns that aren't Year/Month/Day)
  eco_cols <- df %>%
    select(-Year, -Month, -Day, -Variable) %>%
    select(where(is.numeric)) %>%
    names()
  
  # Row-wise mean across eco columns → daily mean tmax across landscape
  df <- df %>%
    mutate(daily_mean_tmax = rowMeans(select(., all_of(eco_cols)), na.rm = TRUE))
  
  # Annual mean
  df %>%
    group_by(Year) %>%
    summarise(annual_mean_tmax = mean(daily_mean_tmax, na.rm = TRUE),
              .groups = "drop") %>%
    mutate(model = var_label)
}
 
# ── Decision models (highlighted) ─────────────────────────────────────────────
decision_models <- c(
  "ACCESS-CM2_ssp585",   # Warm/Dry
  "TaiESM1_ssp585",      # Warm/Wet
  "GISS-E2-1-G_ssp370",  # Cool/Dry
  "INM-CM5-0_ssp370"     # Cool/Wet
)
 
decision_labels <- c(
  "ACCESS-CM2_ssp585"  = "Warm/Dry (ACCESS-CM2 ssp585)",
  "TaiESM1_ssp585"     = "Warm/Wet (TaiESM1 ssp585)",
  "GISS-E2-1-G_ssp370" = "Cool/Dry (GISS-E2-1-G ssp370)",
  "INM-CM5-0_ssp370"   = "Cool/Wet (INM-CM5-0 ssp370)"
)
 
decision_colors <- c(
  "Warm/Dry (ACCESS-CM2 ssp585)"   = "#D73027",   # red
  "Warm/Wet (TaiESM1 ssp585)"      = "#FC8D59",   # orange
  "Cool/Dry (GISS-E2-1-G ssp370)"  = "#4575B4",   # blue
  "Cool/Wet (INM-CM5-0 ssp370)"    = "#74ADD1"    # light blue
)
 
# ── Read future model files ────────────────────────────────────────────────────
csv_files <- list.files(future_dir, pattern = "\\.csv$", full.names = TRUE)
 
future_data <- map_dfr(csv_files, function(fp) {
  # Extract model name from filename, e.g. "ACCESS-CM2_ssp585_NOCA" → "ACCESS-CM2_ssp585"
  model_name <- basename(fp) %>%
    str_remove("_NOCA\\.csv$")
  
  read_annual_tmax(fp, var_label = model_name, tmax_varname = "maxtemp")
})
 
# ── Read historic PRISM ────────────────────────────────────────────────────────
historic_data <- read_annual_tmax(historic_fp, var_label = "Historic PRISM", tmax_varname = "tmax")
 
# ── Combine and classify ───────────────────────────────────────────────────────
# Tag each future model as decision or background
future_data <- future_data %>%
  mutate(
    series_type = if_else(model %in% decision_models, "decision", "background"),
    display_label = case_when(
      model %in% names(decision_labels) ~ decision_labels[model],
      TRUE ~ model
    )
  )
 
historic_data <- historic_data %>%
  mutate(series_type = "historic", display_label = "Historic PRISM (1988–2024)")
 
all_data <- bind_rows(future_data, historic_data)
 
# ── Build plot layers ──────────────────────────────────────────────────────────
 
# Split into background (grey), decision (colored), historic (black dashed)
bg_data       <- filter(all_data, series_type == "background")
decision_data <- filter(all_data, series_type == "decision") %>%
                   mutate(display_label = factor(display_label, levels = names(decision_colors)))
hist_data     <- filter(all_data, series_type == "historic")
 
p <- ggplot() +
  
  # ── Background models (grey, thin) ──────────────────────────────────────────
  geom_line(
    data = bg_data,
    aes(x = Year, y = annual_mean_tmax, group = model),
    color = "grey70", linewidth = 0.35, alpha = 0.6
  ) +
  
  # ── Decision models (colored, bold) ─────────────────────────────────────────
  geom_line(
    data = decision_data,
    aes(x = Year, y = annual_mean_tmax, color = display_label, group = model),
    linewidth = 1.0
  ) +
  
  # ── Historic PRISM (black dashed) ────────────────────────────────────────────
  geom_line(
    data = hist_data,
    aes(x = Year, y = annual_mean_tmax, linetype = display_label),
    color = "black", linewidth = 1.1
  ) +
  
  # ── Scales ───────────────────────────────────────────────────────────────────
  scale_color_manual(
    name   = "Selected Models",
    values = decision_colors
  ) +
  scale_linetype_manual(
    name   = "Observed",
    values = c("Historic PRISM (1988–2024)" = "dashed")
  ) +
  
  # Add a grey dummy entry in the color legend for "Other models"
  # (done via annotate guide hack — simpler approach: add a note in caption)
  
  labs(
    title    = "Annual Mean Daily Maximum Temperature — North Cascades NP (NOCA)",
    subtitle = "All CMIP6 models shown; four decision models highlighted | Landscape-averaged across eco units",
    x        = "Year",
    y        = "Mean Daily Tmax (°C)",
    caption  = "Grey lines = all other CMIP6 models. Decision models: Warm/Dry, Warm/Wet, Cool/Dry, Cool/Wet.\nHistoric: PRISM 800m daily (1988–2024)."
  ) +
  
  theme_bw(base_size = 13) +
  theme(
    plot.title       = element_text(face = "bold", size = 14),
    plot.subtitle    = element_text(size = 10, color = "grey40"),
    legend.position  = "right",
    legend.box       = "vertical",
    legend.key.width = unit(1.5, "cm"),
    panel.grid.minor = element_blank(),
    plot.caption     = element_text(size = 8, color = "grey50", hjust = 0)
  ) +
  
  # Separate legend for the historic line
  guides(
    color    = guide_legend(order = 1, override.aes = list(linewidth = 1.2)),
    linetype = guide_legend(order = 2, override.aes = list(linewidth = 1.2))
  )
 
# ── Save ──────────────────────────────────────────────────────────────────────
out_path <- "~/Dropbox (University of Oregon)/Lamping/NPS_postdoc/Spatial/Climate/NOCA_maxtemp_all_models.png"
 
ggsave(
  filename = out_path,
  plot     = p,
  width    = 14,
  height   = 6,
  dpi      = 300
)
 
message("Figure saved to: ", out_path)
```

# Climate Stream Figure - PPT
```{r}
################################################################################
# plot_NOCA_ppt_all_models.R
#
# Plots annual total precipitation for:
#   - All future climate model CSVs in the NOCA directory
#   - Historic PRISM baseline
#
# Future CSVs: Variable == "ppt", spatial columns eco0:eco22
# Historic CSV: Variable == "ppt", spatial columns eco0:eco22
#
# Four "decision" models are highlighted; all others shown in grey.
################################################################################

library(tidyverse)
library(scales)

# ── Paths ─────────────────────────────────────────────────────────────────────
future_dir  <- "/Users/jlamping/University of Oregon Dropbox/James Lamping/Lamping/NPS_postdoc/Spatial/Climate/outputs/Future_climate_processed/MORA"
historic_fp <- '/Users/jlamping/University of Oregon Dropbox/James Lamping/Lamping/NPS_postdoc/LANDIS/static_landscapes/NOCA/Hist/Rep1/extensions/hist_prism_with_GRIDMET_wind_MORA_03032026.csv'

# ── Helper: read one CSV and return annual total ppt ──────────────────────────
# Averages across all eco columns (landscape mean), then sums across days within each year.
read_annual_ppt <- function(filepath, var_label) {
  message("Reading: ", basename(filepath))
  
  df <- read_csv(filepath, show_col_types = FALSE)
  
  # Keep only rows matching ppt
  df <- df %>% filter(Variable == "ppt")
  
  # Identify eco columns (all numeric columns that aren't Year/Month/Day)
  eco_cols <- df %>%
    select(-Year, -Month, -Day, -Variable) %>%
    select(where(is.numeric)) %>%
    names()
  
  # Row-wise mean across eco columns → daily mean ppt across landscape
  df <- df %>%
    mutate(daily_mean_ppt = rowMeans(select(., all_of(eco_cols)), na.rm = TRUE))
  
  # Annual total (sum of daily values)
  df %>%
    group_by(Year) %>%
    summarise(annual_total_ppt = sum(daily_mean_ppt, na.rm = TRUE),
              .groups = "drop") %>%
    mutate(model = var_label)
}

# ── Decision models (highlighted) ─────────────────────────────────────────────
decision_models <- c(
  "ACCESS-CM2_ssp585",   # Warm/Dry
  "TaiESM1_ssp585",      # Warm/Wet
  "GISS-E2-1-G_ssp370",  # Cool/Dry
  "INM-CM5-0_ssp370"     # Cool/Wet
)

decision_labels <- c(
  "ACCESS-CM2_ssp585"   = "Warm/Dry (ACCESS-CM2 ssp585)",
  "TaiESM1_ssp585"      = "Warm/Wet (TaiESM1 ssp585)",
  "GISS-E2-1-G_ssp370"  = "Cool/Dry (GISS-E2-1-G ssp370)",
  "INM-CM5-0_ssp370"    = "Cool/Wet (INM-CM5-0 ssp370)"
)

decision_colors <- c(
  "Warm/Dry (ACCESS-CM2 ssp585)"   = "#D73027",   # red
  "Warm/Wet (TaiESM1 ssp585)"      = "#FC8D59",   # orange
  "Cool/Dry (GISS-E2-1-G ssp370)"  = "#4575B4",   # blue
  "Cool/Wet (INM-CM5-0 ssp370)"    = "#74ADD1"    # light blue
)

# ── Read future model files ────────────────────────────────────────────────────
csv_files <- list.files(future_dir, pattern = "\\.csv$", full.names = TRUE)

future_data <- map_dfr(csv_files, function(fp) {
  model_name <- basename(fp) %>%
    str_remove("_NOCA\\.csv$")
  
  read_annual_ppt(fp, var_label = model_name)
})

# ── Read historic PRISM ────────────────────────────────────────────────────────
historic_data <- read_annual_ppt(historic_fp, var_label = "Historic PRISM")

# ── Combine and classify ───────────────────────────────────────────────────────
future_data <- future_data %>%
  mutate(
    series_type   = if_else(model %in% decision_models, "decision", "background"),
    display_label = case_when(
      model %in% names(decision_labels) ~ decision_labels[model],
      TRUE ~ model
    )
  )

historic_data <- historic_data %>%
  mutate(series_type = "historic", display_label = "Historic PRISM (1988–2024)")

all_data <- bind_rows(future_data, historic_data)

# ── Build plot layers ──────────────────────────────────────────────────────────
bg_data       <- filter(all_data, series_type == "background")
decision_data <- filter(all_data, series_type == "decision") %>%
                   mutate(display_label = factor(display_label, levels = names(decision_colors)))
hist_data     <- filter(all_data, series_type == "historic")

p <- ggplot() +
  
  # ── Background models (grey, thin) ──────────────────────────────────────────
  geom_line(
    data = bg_data,
    aes(x = Year, y = annual_total_ppt, group = model),
    color = "grey70", linewidth = 0.35, alpha = 0.6
  ) +
  
  # ── Decision models (colored, bold) ─────────────────────────────────────────
  geom_line(
    data = decision_data,
    aes(x = Year, y = annual_total_ppt, color = display_label, group = model),
    linewidth = 1.0
  ) +
  
  # ── Historic PRISM (black dashed) ────────────────────────────────────────────
  geom_line(
    data = hist_data,
    aes(x = Year, y = annual_total_ppt, linetype = display_label),
    color = "black", linewidth = 1.1
  ) +
  
  # ── Scales ───────────────────────────────────────────────────────────────────
  scale_color_manual(
    name   = "Selected Models",
    values = decision_colors
  ) +
  scale_linetype_manual(
    name   = "Observed",
    values = c("Historic PRISM (1988–2024)" = "dashed")
  ) +
  scale_y_continuous(labels = comma) +
  
  labs(
    title    = "Annual Total Precipitation — North Cascades NP (NOCA)",
    subtitle = "All CMIP6 models shown; four decision models highlighted | Landscape-averaged across eco units",
    x        = "Year",
    y        = "Annual Total Precipitation (mm)",
    caption  = "Grey lines = all other CMIP6 models. Decision models: Warm/Dry, Warm/Wet, Cool/Dry, Cool/Wet.\nHistoric: PRISM 800m daily (1988–2024)."
  ) +
  
  theme_bw(base_size = 13) +
  theme(
    plot.title       = element_text(face = "bold", size = 14),
    plot.subtitle    = element_text(size = 10, color = "grey40"),
    legend.position  = "right",
    legend.box       = "vertical",
    legend.key.width = unit(1.5, "cm"),
    panel.grid.minor = element_blank(),
    plot.caption     = element_text(size = 8, color = "grey50", hjust = 0)
  ) +
  
  guides(
    color    = guide_legend(order = 1, override.aes = list(linewidth = 1.2)),
    linetype = guide_legend(order = 2, override.aes = list(linewidth = 1.2))
  )

# ── Save ──────────────────────────────────────────────────────────────────────
out_path <- "~/Dropbox (University of Oregon)/Lamping/NPS_postdoc/Spatial/Climate/NOCA_ppt_all_models.png"

ggsave(
  filename = out_path,
  plot     = p,
  width    = 14,
  height   = 6,
  dpi      = 300
)

message("Figure saved to: ", out_path)
```

