---
title: "Making the ecoregions map Lauren-W"
author: "Lauren Walker"
output: html_notebook
editor_options: 
  chunk_output_type: console
---

This notebook creates the ecoregion map for landis-ii simulations. This script can be used for those using LOCA2 climate change data and making ecoregion map from that raster template. 

```{r}
library(terra)
library(tidyverse)
library(tidyterra)
```


Read in landscape extent and one future climate raster (Source: LOCA2, can be downloaded here: https://cirrus.ucsd.edu/~pierce/LOCA2/)
```{r}
# Define custom projection
albers_custom <- "+proj=aea +lat_1=29.5 +lat_2=45.5 +lat_0=23 +lon_0=-121 +datum=NAD83 +units=m +no_defs"

# Load in landscape polygon perimeter and project
landscape<-vect("Data/Landscape_extent/EAB_landscape_perimeter_new.shp") %>% project(albers_custom)

# Load in one climate file, adjust longitude values, match projection
futureclim<-rast("Data/Climate/Future/ACCESS-CM2/precip/pr.ACCESS-CM2.ssp585.r1i1p1f1.2015-2044.LOCA_16thdeg_v20240915.n_west.monthly.nc") %>% select(pr_tavg_1) %>% project(crs(landscape)) # %>% shift(dx=-360) 

# Export raster to compare with ecoregion in arcgis
#writeRaster(template, "Exports/climatetest.tif")
```


Create ecoregion map from landscape, resample to future climate template
```{r}
# Create template
template <- rast(extent = ext(landscape), crs = crs(landscape), resolution=90)

# Crop the template raster to match landscape extent
futureclim_crop <- crop(futureclim, landscape)

# Assign unique zone IDs to valid template cells (non-NA)
valid_cells <- !is.na(values(futureclim_crop))
values(futureclim_crop)[valid_cells] <- seq_len(sum(valid_cells))

# Resample to 90m using template with nearest neighbor method
ecoregions <- resample(futureclim_crop, template, method = "near")

# Mask to the landscape boundary (optional)
ecoregions <- mask(ecoregions, landscape)

# Reassign values so they go from 1:N (in case some coarse zones are skipped)
unique_vals <- sort(na.omit(unique(values(ecoregions))))
id_map <- setNames(seq_along(unique_vals), unique_vals)
ecoregions[] <- id_map[as.character(values(ecoregions))]

# Plot to check
plot(ecoregions)
```


NEW STEP: aggregate by factor of 200 because there's too many ecoregions.
```{r}
# Aggregate to coarser grid (e.g., 10x10 blocks => 900m)
ecoreg_agg <- aggregate(ecoregions, fact = 200, fun = modal, na.rm = TRUE)

# Relabel the super-ecoregions 1:N
valid_cells <- !is.na(values(ecoreg_agg))
values(ecoreg_agg)[valid_cells] <- seq_len(sum(valid_cells))
plot(ecoreg_agg)

# Resample to 90m 
ecoreg_super <- resample(ecoreg_agg, ecoregions, method = "near")
plot(ecoreg_masked)

# Assign NAs from original ecoregion map to this one
ecoreg_masked <- ifel(is.na(ecoregions), NA, ecoreg_super)

# Fill NAs with zero to make raster a rectangle
ecoreg_masked<- ifel(is.na(ecoreg_masked),0,ecoreg_masked)
```


Now we need to adjust the map for areas where we don't want to grow trees (disturbed, water, etc). TreeMap masks out areas based on LANDFIRE land cover data (Existing Vegetation Type, EVT) which includes human-modified or non-vegetated types from NLCD data.


```{r}
# Read in most recent NLCD land cover raster downloaded from https://www.usgs.gov/centers/eros/science/annual-national-land-cover-database
landcover<-rast("Data/NLCD/Annual_NLCD_LndCov_2023_CU_C1V0_0uhEsq9q3JR5tsws3x2l.tiff")

# Match projections
landcover <- landcover %>% project(crs(ecoregions))

# Resample first to match resolution/extent
landcover_raster <- resample(landcover, ecoregions, method = "near")
```

Our designations:

Ecoregion     Land cover class
0             11 (open water), 12 (perennial snow and ice), 24 (Developed, high                intensity), 31 (barren land)
1             23 (Developed, medium intensity)
2             22 (Developed, low intensity)
3             21 (Developed, open space)
4             81 (hay/pasture), 82 (cultivated crops)


(We might want to exclude a few more categories based on TreeMap if it looks bad!!!)


```{r}
# Step 1: Identify border pixels (original zeros)
border_mask <- ecoreg_masked == 0

# Step 2: Initialize output with ecoregion + 4 (for non-border cells)
out_rast <- ecoreg_masked + 4

# Step 3: Apply land cover rules — only overwrite interior values
out_rast <- ifel(landcover_raster %in% c(11, 12, 24, 31), 0, out_rast)
out_rast <- ifel(landcover_raster == 23, 1, out_rast)
out_rast <- ifel(landcover_raster == 22, 2, out_rast)
out_rast <- ifel(landcover_raster == 21, 3, out_rast)
out_rast <- ifel(landcover_raster %in% c(81, 82), 4, out_rast)

# Step 4: Now restore original zeros
out_rast <- ifel(border_mask, 0, out_rast)

# Export raster
writeRaster(out_rast, "Data/NECN_Final_Maps/EAB_ecoregions_08042025.tif", overwrite= TRUE, datatype = "INT1U")



ecoreg<-rast("Data/NECN_Final_Maps/EAB_ecoregions_08042025.tif")
soilmoist<-rast("C:/Users/lwalk/Desktop/LANDIS_SIMS/Rep6/soilmoist_09082025.tif")

landscape<-vect("Data/Landscape_extent/EAB_landscape_perimeter_new.shp")


```















### Experimental ecoregion map with one ecoregion

Read in landscape extent
```{r}
# Define custom projection
albers_custom <- "+proj=aea +lat_1=29.5 +lat_2=45.5 +lat_0=23 +lon_0=-121 +datum=NAD83 +units=m +no_defs"

# Load in landscape polygon perimeter and project
landscape<-vect("Data/Landscape_extent/EAB_landscape_perimeter.shp") %>% project(albers_custom)

# Create template raster from landscape
template <- rast(extent = ext(landscape), crs = crs(landscape), resolution=90)

# Values to 1
r <- rasterize(landscape, template, field = 1)

# Fill NAs with zero to make raster a rectangle
ecoregion_one <- ifel(is.na(r),0,r)

plot(ecoregion_one)

writeRaster(ecoregion_one, "Data/NECN_Final_Maps/ecoregion_one.tif", datatype= "INT1U")
```
